### 2017-04-13 paper/oopsla17-submission/abstract.tex
asd

\begin{abstract}
  Localizing type errors is challenging in languages with global type
  inference, as the type checker must make assumptions about what the
  programmer intended to do.
  %

  We present a data-driven approach to type error localization that uses
  machine learning to assign blame based on previous
  ill-typed programs and their fixes.
  %
  First, we gather a large set of over 4,500 ill-typed student programs
  and their subsequent fixes.
  %
  Second, we frame type error localization as a supervised learning
  problem: to predict, for each expression in the
  program, whether it should be blamed.
  %
  We represent each expression by its syntactic class and type and those
  of its parent and children, associating blame with expressions that
  were changed in the human-fixed program.
  %
  Finally, we train and evaluate a series of off-the-shelf classifiers
  on our suite of programs,
  and find that they improve localization accuracy from a state-of-the-art
  56\% to 72\%.
\end{abstract}

>>>>

\begin{abstract}
Localizing type errors is challenging in
languages with global type inference, as
the type checker must make assumptions
about what the programmer intended to do.
%
We introduce \toolname, a \emph{data-driven}
approach to error localization based on
supervised learning.
%
\toolname works by using a large corpus
of training data --- pairs of ill-typed
programs and their ``fixed'' versions ---
to automatically \emph{learn a model}
of where the error is most likely.
%
When given a new ill-typed program,
\toolname executes the model to
generate a list of likely error
locations ranked by likelihood.
%
We evaluate \toolname by comparing its
precision against the state-of-the-art
on a set of over 4,500 ill-typed Ocaml
programs drawn from introductory
programming classes.
%
We show that when the top-ranked location
is considered, \toolname's data-driven
model is able to correctly predict
the exact sub-expression that needs to
be changed, 74\% of the time, which is
29 points higher than \ocaml, and
18 points higher than the state-of-the-art
\sherrloc tool.
%
Further, \toolname's accuracy surpasses
85\% when we consider the \emph{top-2}
locations and reaches 91\% if consider
the \emph{top-3}.
\end{abstract}

### 2017-04-13 paper/oopsla17-submission/conclusion.tex
asd

>>>>

We have presented \toolname which
combining modern statistical methods
with domain-specific feature engineering
to open the door to a new data-driven
path to precise error localization.
%
\toolname
translates pairs of
blame-labeled ill-typed
programs into bags-of-abstracted-terms,
and then trains a classifier that
predicts the value of the
whether a feature vector gets
a blame label.
%
Given a new ill-typed program,
we extract its bag-of-extracted-terms
to compute the blame label for each
term and rank the predictions based
on the model's confidence, to select
the top 1-3 blame assignments to
present to the user.

We have evaluated our approach with
five different classifiers on a suite
of over 4,500 student programs.
We find that while machine
learning on its own performs slightly worse
than existing purely constraint based
approaches (\eg \ocaml, \sherrloc), once
we augment the data with feature abstraction
corresponding to the \emph{type error slice}~\cite{Tip2001-qp},
the resulting models models outperform
the state of the art by up to 18
percentage points.

As with other forms of machine learning,
a key concern is that of \emph{data-set bias}: are
the models learned by \toolname specific
to our data set, and would they fail on
\emph{other} \ocaml programs?
%
We address this concern in two ways.
%
First, we partition the data by year,
and show that models learned from one
year generalize to, \ie perform nearly
as well on the programs from the other
year.
%
Second, we argue that in our setting
this bias is \emph{feature} (and not
a bug): it allows \toolname to \emph{adapt}
to the kinds of errors that programmers
(specifically, novices who are in greatest
need of precise feedback) actually make
rather than hardwiring the biases of
compiler authors who, by dint of their
training and experience, may suffer from
blind spots with regards to such problems.
%
In this regard, we are particularly pleased
that our classifiers can be trained on a
modest amount of data, \eg a single course's
worth, and envision a future where each course
has a model of its students' errors.

### 2017-04-13 paper/oopsla17-submission/intro2.tex
asd

In Robin Milner's memorable words
%
``types are the leaven of computer programming;
  they make it digestible.''
%
Languages like \ocaml and \haskell make
the value-proposition for types even more
appealing by using constraints to (mostly)
automatically synthesize the types for all
program terms without troubling the
programmer for any annotations.

Unfortunately, this automation has come at a price.
Verbose type annotations (as in, \eg Java) signify
the programmer's \emph{intent} and help to correctly
pinpoint the error when the code is ill-typed.
%
In contrast, lacking such signifiers, automatic
type inference algorithms are prone to report
type errors quite far from their actual source.
%
While this can seem like a minor annoyance to
veteran programmers, it turns out to be a significant
stumbling block for novices~\citep{Wand1986-nw,Joosten1993-yx}.

@@

This minimum-weight approach suffers
from several drawbacks.
%
First, they are not \emph{portable}:
the algorithms for computing
the minimum weighted subset must be
designed afresh for different kinds
of type systems and constraints.
%
Second, they are not \emph{extensible}:
the weights and analyses are limited
to attributes of the constraints, and
not other \emph{signals} from the
program's syntax that may be more
germane to error localization.
%
Third, and perhaps most importantly,
they are not \emph{adaptable}: the
actual ``weights'' are fixed in
an ad-hoc fashion, based on the
\emph{analysis designers} notion
of what kinds of errors are more
or less likely, rather than
capturing and adapting to the
ground truth about the kinds
of errors programmers actually
make in practice.

\mypara{A Data-Driven Approach}
%
In this paper, we introduce \toolname
a \emph{data-driven} approach to error
localization.
%
\toolname works by using a large corpus
of training data --- pairs of ill-typed
programs and their ``fixed'' versions ---
to automatically \emph{learn a model}
of where the error is most likely.
%
When given a new ill-typed program,
\toolname simply executes the model
to generate a list of likely error
locations ranked by likelihood.
%
We evaluate \toolname by comparing its
precision against the state-of-the-art
on a set of over 4,500 ill-typed Ocaml
programs drawn from introductory
programming classes.
%
We show that \toolname's data-driven
model is able to correctly predict
the exact sub-expression that needs to
be changed, 74\% of the time compared
to \ocaml's 45\% and \sherrloc's 56\%.
%
When the \emph{top} ranked location
is considered. \toolname's accuracy jumps
to above 85\% (resp. 90\%) when we consider
the \emph{top-2} and \emph{top-3} locations.

% - 45
% - 56
% - 74
%
% - 85
% - 91

AST-Dispersion

AST

\toolname addresses both of the limitations
of previous

* portable   [use different features]
* extensible [add more features]
* adaptable  [add more data]

%
Our solution is to \emph{learn} how to
localize type errors by observing
a large set of (novice) errors and their
\emph{subsequent fixes}.
%

\mypara{Contributions}
%
We accomplish this via three concrete contributions.

\paragraph{\textbf{(1) A Dataset of Novice \ocaml Programs}}
Our first contribution is a large dataset of novice \ocaml programs,
including, but not limited to, \emph{ill-typed} programs.
%
We used an instrumented version of the \ocaml compiler
to collect student interactions over two instances of
an undergraduate Programming Languages course at
%
\begin{anonsuppress}
UC San Diego (IRB \#140608).
\end{anonsuppress}
\begin{noanonsuppress}
AUTHOR's INSTITUTION (IRB HIDDEN).
\end{noanonsuppress}
%
The data consist of a time-series of programs submitted to the \ocaml
compiler, from which we extracted a set of \emph{ill-typed} programs
and their \emph{subsequent fixes}.
%
Our dataset comprises over 4,500 such pairs of programs, and many more
programs that are either well-typed or lack a corresponding fix.
%
The entire dataset (both the raw time-series and our extracted pairs of
ill-typed programs and fixes) is available at
%
\begin{anonsuppress}
\ES{TODO: finally post the data somehwere..}.
\end{anonsuppress}
\begin{noanonsuppress}
HIDDEN FOR DOUBLE-BLIND REVIEW.
\end{noanonsuppress}
%

\paragraph{\textbf{(2) Data-Driven Diagnosis of Type Errors}}
Our second contribution is a \emph{data-driven} technique for localizing
type errors (\autoref{sec:learning}).
%
We use machine learning techniques to train a set of classifiers that can
predict, given an ill-typed program, which expressions should be changed
for the program to type check.
%
Our classifiers can be trained on a relatively small set of programs
(\ie a single course's worth).
%
% and outperform the state of the art by
% 15--30\% in blame accuracy.
%
The key technical challenge that we tackle is defining a suitable
interface between the type checker and the classifier, specifically
\emph{embedding} programs as feature vectors and presenting the
classifier's predictions to the user.

\paragraph{\textbf{(3) A Large-Scale Evaluation of Type Error Localization}}
Our third contribution is an evaluation of our classifiers and the
state-of-the-art techniques on our suite of over 4,500 student programs
(\autoref{sec:evaluation}), the largest evaluation of type error
localization techniques we are aware of.
%
We judge the correctness of a localization by comparing it to the
student's eventual fix and find that our classifiers are 15--30\% more
accurate than the state of the art.
%
We also examine which features contribute the most to our
classifiers' predictions, and endeavor to \emph{explain} specific
predictions that our classifiers get right (and wrong).

Overall, our results show that data-driven diagnosis is a practical
technique for localizing type errors by learning from past mistakes.

%
% We believe, furthermore, that with the right set of features our
% approach could be applied to other domains where there are multiple,
% equally valid explanations for an error.
% \ES{this sounds really dumb... need better phrasing}

Consider the |sumList| program in \autoref{fig:sumList},
written by a student in an undergraduate Programming Languages
course.
%
The program is meant to sum the integers in a list, but
the student has accidentally given the empty list as the
base case, rather than 0.
%
The \ocaml compiler collects typing constraints as it
traverses the program, and reports an error the moment
it finds an inconsistent constraint.
%
In this case it blames the recursive call to |sumList|,
complaining that |sumList| returns a |list| where an |int|
was expected by the |+| operator.
%
Note that this \emph{blame} is inconsistent with the
programmer's intention and may not be helpful debugging
information to a novice.

It may appear obvious to the reader that
|[]| is the correct expression to blame,
but how is the type checker to know that?

Indeed, both the venerable
Ocaml compiler as well as
recent techniques~\citep{Zhang2014-lv,Loncaric2016-uk,Pavlinovic2014-mr}.
fail to distinguish between
the two underlined expressions
in \autoref{fig:sumList};
it would be equally valid
to blame \emph{either}
of them alone.
%
The |[]| on line 3 could be changed to |0|,
or the |+| on line 4 could be changed to
either |@| (list append) or |::|, all of
which would give type-correct programs.
%
Thus, these state-of-the-art techniques
are forced to either blame \emph{both}
locations, or choose one \emph{arbitrarily}.

\begin{figure}[ht]
\begin{minipage}{0.45\linewidth}
\begin{ecode}
  let rec sumList xs =
    match xs with
    | []   -> (*@\hlTree{\hlSherrloc{[]}}@*)
    | h::t -> h (*@\hlTree{+}@*) (*@\hlFix{sumList t}@*)
\end{ecode}
\end{minipage}
\begin{minipage}{0.49\linewidth}
\begin{verbatim}
File "sumList.ml", line 4, characters 16-25:
  This expression has type 'a list
  but an expression was expected of type int
\end{verbatim}
\end{minipage}
\caption{(left) An ill-typed \ocaml program that should sum the elements of a
  list, with highlights indicating three possible blame assignments based on:
  %
  (1) the \hlFix{\ocaml} compiler;
  %
  (2) the \hlSherrloc{fix} made by the programmer; and
  %
  (3) \hlTree{minimizing} the number of edits required.
  % The \underline{underlined} expressions are equally valid
  % locations to blame. The expression blamed by the \ocaml compiler
  % is in \textbf{bold}.
  %
  % FIXME: This bolding is ambiguous, because ``let rec'', ``match'' and
  % ``with'' are also bolded (by \\ecode)! You need to find another way to
  % highlight what ocaml is yelling about.
  %
  (right) The error reported by \ocaml.}
\label{fig:sumList}
\end{figure}

% \begin{enumerate}
% \item An large-scale evaluation of state-of-the-art type error
%   localization techniques on over 4,500 student programs.
%   %
%   We collected a time-series of \ocaml programs from our students, and
%   use their subsequent \emph{fixes} to ill-typed programs as
%   \emph{oracles} for the location of the error.
%   %
%   This is the most extensive evaluation of the accuracy of type error
%   localization techniques that we are aware of.
%   % based not on expert \emph{opinion}, but
%   % on the eventual \emph{fix} implemented by a novice user.
%   %
%   % \ES{need to explain \emph{why} this is superior..}
% \item A machine learning approach to \emph{modeling} (novice) type
%   errors that can outperform the state-of-the-art by 15--30\% in
%   localizing errors.
%   %
%   Our classifiers can be trained on a relatively small amount of data --- we
%   used a single class of around 50 students --- and appear to
%   generalize to other instances of the same class.
%   %
%   \ES{and hopefully completely separate classes!}
%   %
%   In contrast to many approaches to \emph{fault localization} from the
%   software engineering community, our model does not use any linguistic
%   modeling techniques, \eg \emph{n-grams} over the token stream; rather,
%   it relies entirely on features of the abstract syntax tree and a
%   partial typing derivation.
%   %
%   \ES{if there's time, investigate adding Wes' n-gram model}
% \end{enumerate}

>>>>

% In Robin Milner's memorable words
% %
% ``types are the leaven of computer programming;
  % they make it digestible''~\cite{tapl}.
%
Types are wonderful.
%
Languages like \ocaml and \haskell make
the value-proposition for types even more
appealing by using constraints to automatically
synthesize the types for all program terms
without troubling the programmer for any
annotations.
%
Unfortunately, this automation has come at a price.
Verbose type annotations (as in, \eg, Java) signify
the programmer's intent and help to correctly
blame the erroneous sub-term when the code is
ill-typed.
%
In contrast, lacking such signifiers, automatic
type inference algorithms are prone to report
type errors quite far from their actual source.
%
While this can seem like a minor annoyance to
veteran programmers, it turns out to be a significant
stumbling block for novices~\citep{Wand1986-nw,Joosten1993-yx}.

@@

This minimum-weight approach suffers
from several drawbacks.
%
First, they are not \emph{extensible}:
the weights and analyses are limited
to attributes of the constraints, and
not other \emph{signals} from the
program's syntax that may be more
germane to error localization. Furthermore,
the algorithms for computing
the minimum weighted subset must be
designed afresh for different kinds
of type systems and constraints.
%
Second, and perhaps most importantly,
they are not \emph{adaptable}: the
weights are fixed in an ad-hoc fashion, based on the
\emph{analysis designers} notion
of what kinds of errors are more
or less likely, rather than
capturing and adapting to the
ground truth about the kinds
of errors programmers actually
make in practice.

\mypara{A Data-Driven Approach}
%
In this paper, we introduce \toolname \footnote{Numeric Analysis of Type Errors}
a \emph{data-driven} approach to error
localization based on supervised
classification~\citep{Kotsiantis2007-pj}.
%
\toolname works by using a large corpus
of training data --- pairs of ill-typed
programs and their ``fixed'' versions ---
to automatically \emph{learn a model}
of where the error is most likely.
%
When given a new ill-typed program,
\toolname simply executes the model
to generate a list of likely error
locations ranked by likelihood.
%
We evaluate \toolname by comparing its
precision against the state-of-the-art
on a set of over 4,500 ill-typed Ocaml
programs drawn from introductory
programming classes.
%
We show that \toolname's data-driven
model is able to correctly predict
the exact sub-expression that needs to
be changed, 74\% of the time compared
to \ocaml's 45\% and \sherrloc's 56\%.
%
When the \emph{top} ranked location
is considered. \toolname's accuracy jumps
to above 85\% (resp. 90\%) when we consider
the \emph{top-2} and \emph{top-3} locations.
%
We achieve these advances by identifying
and then solving three key challenges.

\mypara{Challenge 1: Acquiring Labeled Programs}
%
The first challenge for supervised learning
is to acquire a corpus of training data, here,
a set of ill-typed programs \emph{labeled}
with the exact sub-terms that are the actual
cause of the type error.
%
Prior work has often enlisted expert users
to manually judge ill-typed programs and
determine the ``correct'' fix \citep[\eg][]{Loncaric2016-uk},
but this approach does not scale well to
a dataset large enough to support machine
learning.
%
Worse, while expert users have intimate
knowledge of the type system, they may
have a blind spot with regards to the
kinds of mistakes novices make, and
cannot know in general what novice users
intended to do.

Our \emph{first contribution} \autoref{sec:overview}
is a set of more than 4,500 labeled programs,
gives us an accurate ground truth of
the kinds of errors and the (locations
of the) fixes that novices make in
practice.
%
We obtain this set by observing that
software development is an iterative
process and programmers eventually
fix their own ill-typed programs,
perhaps after multiple incorrect
exploratory attempts.
%
To exploit this observation we instrumented
the \ocaml compiler to collect fine-grained
traces of student interactions over two instances
of an undergraduate Programming Languages course at
%
\begin{anonsuppress}
UC San Diego (IRB \#140608).
\end{anonsuppress}
\begin{noanonsuppress}
AUTHOR's INSTITUTION (IRB HIDDEN).
\end{noanonsuppress}
%
We can then post-process the resulting time-series
of programs submitted to the \ocaml compiler into
a set of pairs of ill-typed programs and their
\emph{fixed} versions, the \emph{first} (type-)correct
program in the trace suffix.
%
We obtain the location labels by computing a
\emph{tree-diff} between the two terms to find
the exact sub-terms that were changed in the fix.

\mypara{Challenge 2: Modeling Programs as Vectors}
%
Modern supervised learning algorithms work on sets of
\emph{feature vectors}: real-valued points in an
$n$-dimensional space. While it is easy to compute such
vectors for documents, images and sound (respectively
word-counts, pixel-values and frequencies), it has
hitherto been unclear how to precisely model program
structure as numeric data.

Our \emph{second contribution} (\autoref{sec:learning})
solves this problem with a new representation called
a \emph{Bag-of-Abstract-Terms (BOAT)} wherein
%
(1)~each program is represented by the \emph{bag}
    or multiset of (sub-)terms that appears inside
    it, and
%
(2)~each (sub-)term $t$ is \emph{abstracted} as
    a vector $[f_1(t), \ldots, f_n(t)]$ where
    each $f_i$ is a \emph{feature abstraction}
    that maps a term to a numeric value.
%
We can model the error labels with a simple feature
abstraction (\eg is-a-diff-subterm).
%
We can even account for \emph{contextual} information like
the features of the parent- and children- terms by
concatenating the feature vectors of each term with those
of its parent and children.

\mypara{Challenge 3: Training Precise Classifiers}
%
Finally, the last and most important challenge is to
use the labeled BOAT data to train classifiers that
are capable of \emph{precisely} pinpointing the error
locations in real programs.
%
The key here is to find the right set of feature
abstractions and classification algorithms, that
lead to precise predictions.
%
Fortunately, our BOAT model allows us a great deal of
latitude in our choice of features.
%
We can use abstraction functions to capture different
aspects of a term ranging from
%
syntactic features (\eg is-a-data constructor, is-a-literal,
is-an-arithmetic-operation, is-a-function-application \etc),
%
to semantic features captured by the type system (\eg is-a-list,
is-an-integer, is-a-function \etc).

Our \emph{third contribution} \autoref{sec:evaluation}
is a systematic evaluation of our data-driven approach
using different classes of features like the above, and
with three different classification algorithms: linear
classifiers, decision forests and deep neural networks.
%
Our evaluation lets us identify the best features and
classifier and empirically characterize the importance
of different features for error localization.
%
In particular, we find that while machine learning
on its own performs slightly worse than existing
purely constraint based approaches (\eg \ocaml, \sherrloc),
once we augment the data with feature abstraction
corresponding to the \emph{type error slice} \cite{Tip2001-qp},
the resulting classifier significantly outperforms
state-of-the-art by 18 percentage points.

Thus, by combining modern statistical methods
with domain-specific feature engineering
to open the door to a new data-driven
path to precise error localization.
%
We can easily \emph{extend} \toolname to
new languages or forms of correctness
checks by swapping in a different
set of feature abstraction functions.
%
Furthermore, our data-driven approach
allows \toolname to \emph{adapt} to
the kinds of errors that programmers
(specifically, novices who are in greatest
need of precise feedback) actually make
rather than hardwiring the biases of
compiler authors who, by dint of their
training and experience, may suffer from
blind spots with regards to such problems.
%
In contrast, our results show that \toolname's
data-driven diagnosis can be an effective
technique for localizing errors by collectively
learning from past mistakes.

### 2017-04-13 paper/oopsla17-submission/scratch.tex
asd

\section{Introduction}
\label{sec:introduction}

Type systems are one of the most widely-used techniques for program
verification. A type checker only accepts a program if it can
\emph{prove} the absence of an entire class of bugs.
%
Typed functional languages like \ocaml and \haskell are equipped with
particularly powerful type checkers that can operate without a single
annotation, alleviating a prime concern with verification techniques.
%
Yet this power comes with a cost: type annotations signify the
programmer's \emph{intent}, without them the checker must
make assumptions about what the programmer intended to do.
%
This results in the common phenomenon of type errors being reported far
from their source, an annoyance to veterans of the type system, but a
significant hurdle for novices \citep{Wand1986-nw,Joosten1993-yx}.

\begin{figure}[ht]
\begin{minipage}{0.45\linewidth}
\begin{ecode}
  let rec sumList xs =
    match xs with
    | []   -> (*@\hlTree{\hlSherrloc{[]}}@*)
    | h::t -> h (*@\hlTree{+}@*) (*@\hlFix{sumList t}@*)
\end{ecode}
\end{minipage}
\begin{minipage}{0.49\linewidth}
\begin{verbatim}
File "sumList.ml", line 4, characters 16-25:
  This expression has type 'a list
  but an expression was expected of type int
\end{verbatim}
\end{minipage}
\caption{(left) An ill-typed \ocaml program that should sum the elements of a
  list, with highlights indicating three possible blame assignments based on:
  %
  (1) the \hlFix{\ocaml} compiler;
  %
  (2) the \hlSherrloc{fix} made by the programmer; and
  %
  (3) \hlTree{minimizing} the number of edits required.
  % The \underline{underlined} expressions are equally valid
  % locations to blame. The expression blamed by the \ocaml compiler
  % is in \textbf{bold}.
  %
  % FIXME: This bolding is ambiguous, because ``let rec'', ``match'' and
  % ``with'' are also bolded (by \\ecode)! You need to find another way to
  % highlight what ocaml is yelling about.
  %
  (right) The error reported by \ocaml.}
\label{fig:sumList}
\end{figure}
%
Consider the |sumList| program in \autoref{fig:sumList}, written by a
student in an undergraduate Programming Languages course.
%
The program is meant to sum the integers in a list, but the student has
accidentally given the empty list as the base case, rather than 0.
%
The \ocaml compiler collects typing constraints as it traverses the
program, and reports an error the moment it finds an inconsistent
constraint.
%
In this case it blames the recursive call to |sumList|, complaining that
|sumList| returns a |list| where an |int| was expected by the |+|
operator.
%
Note that this blame is inconsistent with the programmer's intention and
may not be helpful debugging information to a novice.

Recent work on type error \emph{localization}
%
\citep[\eg][]{Zhang2014-lv,Loncaric2016-uk,Pavlinovic2014-mr},
%
determining which expression(s) to blame for the error,
frames localization as an optimization problem, seeking a set of
expressions to blame that minimize some cost function (\eg the number of
constraints in the blame set, or the ratio of ``good'' paths through the
constraint graph to ``bad'' paths).
%
However, even these techniques fail to distinguish between the two
underlined expressions in \autoref{fig:sumList}, it would be equally
valid to blame \emph{either} of them alone.
%
The |[]| on line 3 could be changed to |0|, or the |+| on line 4 could
be changed to either |@| (list append) or |::|, all of which would give
type-correct programs.
%
Thus, these state-of-the-art techniques are forced to either blame
\emph{both} locations, or choose one \emph{arbitrarily}.

It may appear obvious to the reader that |[]| is the correct expression
to blame, but how is the type checker to know that?
%
Our solution is to \emph{learn} how to localize type errors by observing
a large set of (novice) errors and their \emph{subsequent fixes}.
%
We accomplish this via three concrete contributions.

\paragraph{\textbf{(1) A Dataset of Novice \ocaml Programs}}
Our first contribution is a large dataset of novice \ocaml programs,
including, but not limited to, \emph{ill-typed} programs.
%
We used an instrumented version of the \ocaml compiler to collect
student interactions over two instances of an undergraduate Programming
Languages course at
%
\begin{anonsuppress}
UC San Diego (IRB \#140608).
\end{anonsuppress}
\begin{noanonsuppress}
AUTHOR's INSTITUTION (IRB HIDDEN).
\end{noanonsuppress}
%
The data consist of a time-series of programs submitted to the \ocaml
compiler, from which we extracted a set of \emph{ill-typed} programs
and their \emph{subsequent fixes}.
%
Our dataset comprises over 4,500 such pairs of programs, and many more
programs that are either well-typed or lack a corresponding fix.
%
The entire dataset (both the raw time-series and our extracted pairs of
ill-typed programs and fixes) is available at
%
\begin{anonsuppress}
\ES{TODO: finally post the data somehwere..}.
\end{anonsuppress}
\begin{noanonsuppress}
HIDDEN FOR DOUBLE-BLIND REVIEW.
\end{noanonsuppress}
%

\paragraph{\textbf{(2) Data-Driven Diagnosis of Type Errors}}
Our second contribution is a \emph{data-driven} technique for localizing
type errors (\autoref{sec:learning}).
%
We use machine learning techniques to train a set of classifiers that can
predict, given an ill-typed program, which expressions should be changed
for the program to type check.
%
Our classifiers can be trained on a relatively small set of programs
(\ie a single course's worth). 
% WRW notes that we say this below in contribution (3), which is only three
% sentences away, so to a first-time reader it sounds like we are
% double-counting:
%
% and outperform the state of the art by
% 15--30\% in blame accuracy.
%
The key technical challenge that we tackle is defining a suitable
interface between the type checker and the classifier, specifically
\emph{embedding} programs as feature vectors and presenting the
classifier's predictions to the user.

\paragraph{\textbf{(3) A Large-Scale Evaluation of Type Error Localization}}
Our third contribution is an evaluation of our classifiers and the
state-of-the-art techniques on our suite of over 4,500 student programs
(\autoref{sec:evaluation}), the largest evaluation of type error
localization techniques we are aware of.
%
We judge the correctness of a localization by comparing it to the
student's eventual fix and find that our classifiers are 15--30\% more
accurate than the state of the art.
%
We also examine which features contribute the most to our
classifiers' predictions, and endeavor to \emph{explain} specific
predictions that our classifiers get right (and wrong).

Overall, our results show that data-driven diagnosis is a practical
technique for localizing type errors by learning from past mistakes.
%
% We believe, furthermore, that with the right set of features our
% approach could be applied to other domains where there are multiple,
% equally valid explanations for an error.
% \ES{this sounds really dumb... need better phrasing}

% \begin{enumerate}
% \item An large-scale evaluation of state-of-the-art type error
%   localization techniques on over 4,500 student programs.
%   %
%   We collected a time-series of \ocaml programs from our students, and
%   use their subsequent \emph{fixes} to ill-typed programs as
%   \emph{oracles} for the location of the error.
%   %
%   This is the most extensive evaluation of the accuracy of type error
%   localization techniques that we are aware of.
%   % based not on expert \emph{opinion}, but
%   % on the eventual \emph{fix} implemented by a novice user.
%   %
%   % \ES{need to explain \emph{why} this is superior..}
% \item A machine learning approach to \emph{modeling} (novice) type
%   errors that can outperform the state-of-the-art by 15--30\% in
%   localizing errors.
%   %
%   Our classifiers can be trained on a relatively small amount of data --- we
%   used a single class of around 50 students --- and appear to
%   generalize to other instances of the same class.
%   %
%   \ES{and hopefully completely separate classes!}
%   %
%   In contrast to many approaches to \emph{fault localization} from the
%   software engineering community, our model does not use any linguistic
%   modeling techniques, \eg \emph{n-grams} over the token stream; rather,
%   it relies entirely on features of the abstract syntax tree and a
%   partial typing derivation.
%   %
%   \ES{if there's time, investigate adding Wes' n-gram model}
% \end{enumerate}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Consider the |sumList| program in \autoref{fig:sumList},
written by a student in an undergraduate Programming Languages
course.
%
The program is meant to sum the integers in a list, but
the student has accidentally given the empty list as the
base case, rather than 0.
%
The \ocaml compiler collects typing constraints as it
traverses the program, and reports an error the moment
it finds an inconsistent constraint.
%
In this case it blames the recursive call to |sumList|,
complaining that |sumList| returns a |list| where an |int|
was expected by the |+| operator.
%
Note that this \emph{blame} is inconsistent with the
programmer's intention and may not be helpful debugging
information to a novice.

It may appear obvious to the reader that
|[]| is the correct expression to blame,
but how is the type checker to know that?

Indeed, both the venerable
Ocaml compiler as well as
recent techniques~\citep{Zhang2014-lv,Loncaric2016-uk,Pavlinovic2014-mr}.
fail to distinguish between
the two underlined expressions
in \autoref{fig:sumList};
it would be equally valid
to blame \emph{either}
of them alone.
%
The |[]| on line 3 could be changed to |0|,
or the |+| on line 4 could be changed to
either |@| (list append) or |::|, all of
which would give type-correct programs.
%
Thus, these state-of-the-art techniques
are forced to either blame \emph{both}
locations, or choose one \emph{arbitrarily}.

\begin{figure}[ht]
\begin{minipage}{0.45\linewidth}
\begin{ecode}
  let rec sumList xs =
    match xs with
    | []   -> (*@\hlTree{\hlSherrloc{[]}}@*)
    | h::t -> h (*@\hlTree{+}@*) (*@\hlFix{sumList t}@*)
\end{ecode}
\end{minipage}
\begin{minipage}{0.49\linewidth}
\begin{verbatim}
File "sumList.ml", line 4, characters 16-25:
  This expression has type 'a list
  but an expression was expected of type int
\end{verbatim}
\end{minipage}
\caption{(left) An ill-typed \ocaml program that should sum the elements of a
  list, with highlights indicating three possible blame assignments based on:
  %
  (1) the \hlFix{\ocaml} compiler;
  %
  (2) the \hlSherrloc{fix} made by the programmer; and
  %
  (3) \hlTree{minimizing} the number of edits required.
  % The \underline{underlined} expressions are equally valid
  % locations to blame. The expression blamed by the \ocaml compiler
  % is in \textbf{bold}.
  %
  % FIXME: This bolding is ambiguous, because ``let rec'', ``match'' and
  % ``with'' are also bolded (by \\ecode)! You need to find another way to
  % highlight what ocaml is yelling about.
  %
  (right) The error reported by \ocaml.}
\label{fig:sumList}
\end{figure}
